---
title: "Assignment 1"
author: "Deepak Chethan"
date: "April 19, 2018"
output: md_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
fig.path='Figs/'
```

Problem Statement
----------------------------
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

-   Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as 𝙽𝙰) </br> date: The date on which the measurement was taken in YYYY-MM-DD format </br> interval: Identifier for the 5-minute interval in which measurement was taken </br> The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

Step 1: Loading and preprocessing the data
------------------------------------------------------------------
	
	
``` {r}
library(ggplot2)

fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = paste0(getwd(), '/repdata%2Fdata%2Factivity.zip'), method = "curl")
unzip("repdata%2Fdata%2Factivity.zip",exdir = "data")
```

Reading the data 
-------------------------
``` {r, echo=TRUE, results="markup"}
data <- read.csv("data/activity.csv")
```

What is the mean total number of steps taken per day ?
-----------------------------------------------------------------------------
```{r ,echo=TRUE, results="markup"}
steps_taken_per_day <- tapply(data$steps,data$date, sum, na.rm=FALSE)
hist(steps_taken_per_day, xlab='Total steps taken per day',ylab='Frequency', breaks=500)
```


What are the mean and medial total number of steps:
---------------------------------------------------------------------------
```{r ,echo=TRUE, results="markup"}

mean_steps <- mean(steps_taken_per_day,na.rm = TRUE)
media_steps <- median(steps_taken_per_day, na.rm = TRUE)
mean_steps
media_steps
```

What is the average daily activity pattern?
-------------------------------------------------------------

```{r ,echo=TRUE, results="markup"}
avg_steps_per_time <- aggregate(x=list(Avg_Steps=data$steps), by=list(interval=data$interval), FUN=mean, na.rm = TRUE)
plot(avg_steps_per_time$interval,avg_steps_per_time$Avg_Steps,xlab = "5 min interval", ylab = "Avg steps", type = "o")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 
----------------------------------------------------------------------------------------------------------------------------------------
```{r echo=TRUE, results="markup"}
max_steps <- which.max(avg_steps_per_time$Avg_Steps)
time_of_max_steps <-  gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", avg_steps_per_time[max_steps,'interval'])
time_of_max_steps
```

#Imputing Missing Values
----------------------------------------
1. Calculate and report the total number of missing values in the dataset

```{r echo=TRUE, results="markup"}
nMissingVals <- length(which(is.na(data$steps)))
nMissingVals
```
2. Devise a strategy for filling in all of the missing values in the dataset.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r echo=TRUE, results="markup"}
library(imputeMissings)
filledData <- impute(data = data)
```

4. Make a histogram of the total number of steps taken each day

```{r echo=TRUE, results="markup"}
steps_taken_per_day <- tapply(filledData$steps,filledData$date, sum, na.rm=FALSE)
hist(steps_taken_per_day, xlab='Total steps taken per day',ylab='Frequency', breaks=500)
```
... and Calculate and report the mean and median

```{r echo=TRUE, results="markup"}
mean_steps <- mean(filledData$steps)
median_steps <- median(filledData$steps)
mean_steps
median_steps
```

#Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r echo=TRUE, results="markup"}
filledData$dayType <- ifelse(as.POSIXlt(filledData$date)$wday %in% c(0,6),'weekend','weekday')

```

2. Make a panel plot containing a time series plot
```{r echo=TRUE, results="markup"}
avg_steps_per_time_filled <- aggregate(steps ~ interval + dayType, data=filledData, mean)
ggplot(avg_steps_per_time_filled, aes(interval,steps))+
	geom_line()+
	facet_grid(dayType ~ .)+
	xlab("5 min interval")+
	ylab("avg num of steps")
```







